---
title: "Central Limit Theorem Illustration"
author: "Stefan Britz"
date: "University of Cape Town"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
```


## This app demonstrates the Central Limit Theorem interactively

#### First select the distribution of $X$ 

```{r}
inputPanel(
  radioButtons("dist", label = 'Select a type of distribution', 
               choices = c("N(0, 1)" = "stand_norm", "N(10, 25)" = "gaus", "Gamma(2, 5)"= "gam", 
                           "U(1, 6)" = "uni", "Beta(0.5, 0.5)" = "bet", "Poisson(5)" = "pois", 
                           "Binomial(8, 0.1)" = "bin"))
)
```




#### Here is the pdf/pmf of the selected variable:
```{r}

dist_plot <- function(type){
  switch(type,
         stand_norm = curve(dnorm(x), -4, 4, lwd = 2, xlab = 'X ~ N(0, 1)', ylab = "", cex.lab = 2),
         gaus = curve(dnorm(x, 10, 5), -10, 30, lwd = 2, xlab = 'X ~ N(10, 25)', ylab = "", cex.lab = 2),
         gam = curve(dgamma(x, 2, 5), 0, 4, lwd = 2, xlab = 'X ~ Gamma(2, 5)', ylab = "", cex.lab = 2),
         bet = curve(dbeta(x, 0.5, 0.5), 0, 1, lwd = 2, xlab = 'X ~ Beta(0.5, 0.5)', ylab = "", cex.lab = 2),
         uni = curve(dunif(x, 1, 6), 0, 7, lwd = 2, xlab = 'X ~ U(1, 6)', ylab = "", cex.lab = 2),
         pois = barplot(dpois(0:20, 5), names.arg = 0:20, cex.lab = 2, xlab = 'X ~ Pois(5)', ylab = ""),
         bin = barplot(dbinom(0:8, 8, 0.1), names.arg = 0:8, cex.lab = 2, xlab = 'X ~ Bin(8, 0.1)', ylab = "")
  )
  
}
renderPlot({dist_plot(input$dist)})

```

#### We will now sample from this distribution and display the distribution of the sample mean: $\bar{X} = \frac{\sum_{i=1}^{n} X_i}{n}$. Select the sample size (n).


```{r}
inputPanel(
  numericInput("sample_size", "Choose a sample size", 100, 1, 1000) 
)
```

### Below is the distribution of the means from 10 000 of these samples

``` {r}


means_dist <- function(type){
  
  
  randvals <<- switch(type,
                      stand_norm = apply(matrix(rnorm(10000*input$sample_size), nrow = input$sample_size), 2, mean),
                      gaus = apply(matrix(rnorm(10000*input$sample_size, 10, 5), nrow = input$sample_size), 2, mean),
                      gam = apply(matrix(rgamma(10000*input$sample_size, 2, 5), nrow = input$sample_size), 2, mean),
                      bet = apply(matrix(rbeta(10000*input$sample_size, 0.5, 0.5), nrow = input$sample_size), 2, mean),
                      uni = apply(matrix(runif(10000*input$sample_size, 1, 6), nrow = input$sample_size), 2, mean),
                      pois = apply(matrix(rpois(10000*input$sample_size, 5), nrow = input$sample_size), 2, mean),
                      bin = apply(matrix(rbinom(10000*input$sample_size, 8, 0.1), nrow = input$sample_size), 2, mean)
  )
  
  hist(randvals, main =  expression(paste("Histogram of ", bar(X))), xlab = expression(bar(X)), 
       col = 'darkgray', border = 'white', freq = FALSE)
}

dens_dist <- function(type){
  switch(type,
         stand_norm = curve(dnorm(x, 0, sqrt(1/input$sample_size)), add = TRUE, col = "blue"),
         gaus = curve(dnorm(x, 10, sqrt(25/input$sample_size)), add = TRUE, col = "blue"),
         gam = curve(dnorm(x, 2/5, sqrt((2/25)/input$sample_size)), add = TRUE, col = "blue"),
         bet = curve(dnorm(x, 0.5, sqrt((1/8)/input$sample_size)), add = TRUE, col = "blue"),
         uni = curve(dnorm(x, 3.5, sqrt((25/12)/input$sample_size)), add = TRUE, col = "blue"),
         pois = curve(dnorm(x, 5, sqrt(5/input$sample_size)), add = TRUE, col = "blue"),
         bin = curve(dnorm(x, 0.8, sqrt(0.72/input$sample_size)), add = TRUE, col = "blue")
  )
}


renderPlot({
  means_dist(input$dist)
  
  if(input$show_norm == TRUE) dens_dist(input$dist)
  
})
```

#### Tick the following box to display the density function of $\bar{X} \sim N\left(\mu, \frac{\sigma^2}{n}\right)$

```{r}
inputPanel(
  checkboxInput("show_norm", "Fit Normal Density?", FALSE)
)
```

#### Using the emperical moments to check the distributional characteristics of $\bar{X}$:

```{r}
library(moments)

inputPanel(
  actionButton(inputId = "calc_moments", label = "Re-calculate for the above sample")
)

moments <- function(){
  plot.new()
  text(0.2, 0.8, paste("Mean = ", round(mean(randvals), 2)), cex = 3)
  text(0.2, 0.2, paste("Variance = ", round(var(randvals), 5)), cex = 3)
  text(0.75, 0.8, paste("Coef of Skewness = ", round(skewness(randvals), 2)), cex = 3)
  text(0.78, 0.2, paste("Excess Kurtosis = ", round(kurtosis(randvals) - 3, 2)), cex = 3)
}

p <- eventReactive(input$calc_moments, {moments()})

renderPlot({
  p()
})

```

