---
title: "Distribution of Sample Statistics"
author: "Stefan Britz"
date: "University of Cape Town"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Let's again start with sampling distribution of the mean:

### $\bar{X} = \frac{\sum_{i=1}^{n} X_i}{n}$


#### First select the distribution of $X$ 


```{r, echo=FALSE}
inputPanel(
    radioButtons("dist", label = 'Select a distribution', 
                 choices = c("N(0, 1)" = "stand_norm", "N(10, 25)" = "gaus", "Gamma(2, 5)"= "gam", "Beta(0.5, 0.5)" = "bet"))
    
)
```


#### This is the distribution of the population:
```{r echo=FALSE, warning=FALSE}

dist_plot <- function(type){
switch(type,
       stand_norm = curve(dnorm(x), -4, 4, ylab = 'f(x)'),
       gaus = curve(dnorm(x, 10, 5), -10, 30, ylab = 'f(x)'),
       gam = curve(dgamma(x, 2, 5), 0, 4, ylab = 'f(x)'),
       bet = curve(dbeta(x, 0.5, 0.5), 0, 1, ylab = 'f(x)'))
  
}
renderPlot({dist_plot(input$dist)})

```

#### We will now sample from this distribution. Select a sample size


```{r, echo=FALSE}
inputPanel(
    numericInput("sample_size", "Choose a sample size", 100, 1, 1000) 

)
```

### Below is the distribution of the means from 10 000 of these samples

```{r echo=FALSE, warning=FALSE}
means_dist <- function(type){
switch(type,
       stand_norm = hist(apply(matrix(rnorm(10000*input$sample_size), nrow = input$sample_size), 2, mean), 
        main =  expression(paste("Histogram of ", bar(X))), xlab = expression(bar(X)), col = 'darkgray', border = 'white', freq = FALSE),
       gaus = hist(apply(matrix(rnorm(10000*input$sample_size, 10, 5), nrow = input$sample_size), 2, mean), 
        main =  expression(paste("Histogram of ", bar(X))), xlab = expression(bar(X)), col = 'darkgray', border = 'white', freq = FALSE),
       gam = hist(apply(matrix(rgamma(10000*input$sample_size, 2, 5), nrow = input$sample_size), 2, mean), 
        main =  expression(paste("Histogram of ", bar(X))), xlab = expression(bar(X)), col = 'darkgray', border = 'white', freq = FALSE),
       bet = hist(apply(matrix(rbeta(10000*input$sample_size, 0.5, 0.5), nrow = input$sample_size), 2, mean), 
        main =  expression(paste("Histogram of ", bar(X))), xlab = expression(bar(X)), col = 'darkgray', border = 'white', freq = FALSE)
)
}

dens_dist <- function(type){
  switch(type,
         stand_norm = curve(dnorm(x, 0, sqrt(1/input$sample_size)), add = TRUE, col = "blue"),
         gaus = curve(dnorm(x, 10, sqrt(25/input$sample_size)), add = TRUE, col = "blue"),
         gam = curve(dnorm(x, 2/5, sqrt((2/25)/input$sample_size)), add = TRUE, col = "blue"),
         bet = curve(dnorm(x, 0.5, sqrt((1/8)/input$sample_size)), add = TRUE, col = "blue")
           )
}


renderPlot({
  means_dist(input$dist)
  
  if(input$show_norm == TRUE) dens_dist(input$dist)
  
  })
```

#### Tick the following box to display the density function of $\bar{X} \sim N\left(\mu, \frac{\sigma^2}{n}\right)$

```{r, echo=FALSE}
inputPanel(
    checkboxInput("show_norm", "Fit Normal Density?", FALSE)

)
```


## Now we will also explore the distribution of the sample Variance: 

### $S^2 = \frac{\sum_{i = 1}^{n}(X_i - \bar{X})^2}{n-1}$

### Here we see the distribution of $Y = \frac{(n-1)S^2}{\sigma^2}$ from 10 000 of these samples:

```{r echo=FALSE, warning=FALSE}
vars_dist <- function(type){
switch(type,
       stand_norm = hist((input$sample_size - 1)/1*apply(matrix(rnorm(10000*input$sample_size), nrow = input$sample_size), 2, var), 
        main = expression(paste("Histogram of Y = ", frac((n-1)*S^2, sigma^2))), xlab = "Y", col = 'darkgray', border = 'white', freq = FALSE),
       gaus = hist((input$sample_size - 1)/25*apply(matrix(rnorm(10000*input$sample_size, 10, 5), nrow = input$sample_size), 2, var), 
        main = expression(paste("Histogram of Y = ", frac((n-1)*S^2, sigma^2))), xlab = "Y", col = 'darkgray', border = 'white', freq = FALSE),
       gam = hist((input$sample_size - 1)/(2/25)*apply(matrix(rgamma(10000*input$sample_size, 2, 5), nrow = input$sample_size), 2, var), 
        main = expression(paste("Histogram of Y = ", frac((n-1)*S^2, sigma^2))), xlab = "Y", col = 'darkgray', border = 'white', freq = FALSE),
       bet = hist((input$sample_size - 1)/(1/8)*apply(matrix(rbeta(10000*input$sample_size, 0.5, 0.5), nrow = input$sample_size), 2, var), 
        main = expression(paste("Histogram of Y = ", frac((n-1)*S^2, sigma^2))), xlab = "Y", col = 'darkgray', border = 'white', freq = FALSE)
)
}

dens_dist2 <- function(type){
  switch(type,
         stand_norm = curve(dchisq(x, input$sample_size - 1), add = TRUE, col = "blue"),
         gaus = curve(dchisq(x, input$sample_size - 1), add = TRUE, col = "blue"),
         gam = curve(dchisq(x, input$sample_size - 1), add = TRUE, col = "blue"),
         bet = curve(dchisq(x, input$sample_size - 1), add = TRUE, col = "blue")
           )
}


renderPlot({
  vars_dist(input$dist)
  
  if(input$show_chi == TRUE) dens_dist2(input$dist)
  
  })
```

#### Tick the following box to display the density function of $\chi^2_{n-1}$

```{r, echo=FALSE}
inputPanel(
    checkboxInput("show_chi", "Fit Density?", FALSE)

)
```